A Plasma Torus Around a Young Low-Mass Star

### 2025-03-12 paper/ms.tex
update

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
phenomenologically identified based on their highly structured and
periodic optical light curves;
most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.png}
  \caption[]{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).} For the best experience, please view the online movie
  available
  \href{https://lgbouma.com/movies/movie_TIC1411_flux_phase.mp4}{here},
  which spans a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period in
  groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute
  sampling are in gray; black is their average.  Similar to other members
  of this class, the sharp photometric features persist for tens to
  thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

@@

In this study, we present the first observations of corotating clumps
of cool plasma around a CPV.  We identified TIC~141146667 in previous
work \cite{Bouma2024} by searching the TESS two-minute data for stars
showing periodic variability with at least three sharp dips per cycle.
We selected the star for spectroscopic observations because its
brightness and rapid rotation offered sensitivity to small variations
in the line profiles.  We observed for five hours on UT 2024-02-17
using the High Resolution Echelle Spectrometer (HIRES;
\cite{vogt_hires_1994}) on the Keck I 10m telescope.  TESS observed
the star from UT 2024-02-05 to UT 2024-02-26 with a duty cycle of
XX\%.  
At the exact instant of spectral observations, TESS was performing a
data downlink; photometric data collection resumed three rotation
cycles (12 hours) after the spectra were acquired.  Extended Data
Figure~\ref{fig:fulllc} shows the detailed photometric behavior of the
star before and after the exact epoch of observation; the photometric
morphology is similar before and after the data gap.

@@

1: SED fit\cite{Bouma2024},
2: Gaia DR3 empirical isochrone fit (see~\ref{subsec:stparams}),
3: PARSEC %(TODO)

@@

5: TESS light curve,
6: Derived quantity,

>>>>

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main object of
interest in this article, over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
phenomenologically identified based on their highly structured and
periodic optical light curves;
most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

@@

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.png}
  \caption[]{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).} For the best experience, please view the online movie
  available
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{here},
  which spans a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period in
  groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute
  sampling are in gray; black is their average.  Similar to other members
  of this class, the sharp photometric features persist for tens to
  thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

@@

In this study, we present the first observations of corotating clumps of
cool plasma around a CPV.  We identified TIC~141146667 in previous work
\cite{Bouma2024} by searching the TESS two-minute data \cite{Ricker2015}
for stars showing at least three sharp and periodic dips per cycle.  We
selected the star for spectroscopy because its brightness and rapid
rotation offer sensitivity to small variations in the line profiles.  We
observed for five hours on 17~February~2024 (UT) using the High
Resolution Echelle Spectrometer (HIRES; \cite{vogt_hires_1994}) on the
Keck I 10m telescope.  TESS observed the star over a four week period
from 30~January~2024 to 26~February~2024 with a duty cycle of 77\%.
During the spectral observations, TESS was performing a data downlink;
TESS data collection resumed 12 hours (three rotation cycles) after the
spectra were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star before and after the exact
epoch of observation; the photometric morphology is similar before and
after the data gap.

@@

1: SED fit \cite{Bouma2024}.
2: Gaia DR3 empirical isochrone fit (see~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.

@@

5: TESS light curve.
6: Derived quantity.
